\begin{center}
{\Large \textbf{Abstract}}
\end{center}
Recent progress in natural language processing has been largely driven by large-scale neural 
networks. These architectures aim to learn representations of text that capture syntax, 
semantics, and long-range dependencies in an efficient and scalable way. 
Despite significant advances in performance, the internal mechanisms by which these models store and manipulate information 
remain only partially understood. \\ \\
% 
This limitation becomes especially relevant in modern architectures that rely on fundamentally 
different approaches to information storage. Transformer-based models process sequences using 
self-attention, where information is mixed across tokens at each layer without an explicit notion 
of persistent state. In contrast, xLSTM introduced by Beck et al. \cite{beck2024xlstmextendedlongshortterm} 
a modern recurrent neural network architecture represents an alternative to the transformer based architectures 
by incorporating explicit memory that is updated incrementally over time. This allows information to be stored 
and refined accross consecutive time steps. \\ \\ 
%
This work investigates the internal behavior of the xLSTM architecture through a set of interpretability 
experiments aimed at understanding how information is represented, stored, and transformed inside the 
model. The analysis explores multiple levels of abstraction, from token-level dynamics to layer-wise 
representation structure. Specifically, we apply logit lens\cite{nostalgebraist2020logitlens} 
\cite{belrose2023tunedlens} to observe the relationship between the internal representations in the embedding 
space and their mapping into the vocabulary space. Moreover, we compare the hidden states of xLSTM and a comparable 
Transformer-based model (Mistral 7B) \cite{jiang2023mistral7b} to assess how similarly semantic representations evolve 
in the two models. Furthermore, we quantify memory behavior by tracking the 
evolution of the cell state over time and tracking input and forget gate activations at the 
token level to understand how the model selectively retains or discards information. 
Together, these experiments provide a multi-perspective view into the internal mechanisms of modern recurrent 
neural networks, offering interpretable details about how structured memory evolves in large-scale recurrent architectures.\\ \\
%
Finally, we investigate a practical application of xLSTM's fixed-size memory state by caching contextual 
information for question-answering tasks. This approach enables inference speedups that increase with context
length, as the cached memory can be reused without reprocessing the entire input sequence. 
We evaluate the method against Mistral 7B and analyze the trade-offs between recurrent and Transformer-based 
architectures with respect to inference speed and task performance.


